---
title: "Final Project: Time Series Analysis of GDP, Exports, and Population"
author: "Andrew"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup

## Load Data

```{r load-data}
load("/Users/miguelflores/Downloads/Data_File_finalproject.Rdata")
```

## Load Libraries

```{r load-libraries}
library(forecast)
library(astsa)
library(lmtest)
library(ggplot2)
library(dplyr)
library(gridExtra)
```

## Prepare Data Frame

```{r prep-df}
df <- as.data.frame(finalPro_data)
```

# Exploratory Analysis

## Summary Statistics — Central African Republic

```{r summary-stats}
finalPro_data %>%
  filter(Country == "Central African Republic") %>%
  summarise(
    Mean = c(mean(GDP, na.rm = TRUE), mean(Growth, na.rm = TRUE),
             mean(CPI, na.rm = TRUE), mean(Imports, na.rm = TRUE),
             mean(Exports, na.rm = TRUE), mean(Population, na.rm = TRUE)),
    Median = c(median(GDP, na.rm = TRUE), median(Growth, na.rm = TRUE),
               median(CPI, na.rm = TRUE), median(Imports, na.rm = TRUE),
               median(Exports, na.rm = TRUE), median(Population, na.rm = TRUE)),
    SD = c(sd(GDP, na.rm = TRUE), sd(Growth, na.rm = TRUE),
           sd(CPI, na.rm = TRUE), sd(Imports, na.rm = TRUE),
           sd(Exports, na.rm = TRUE), sd(Population, na.rm = TRUE)),
    Min = c(min(GDP, na.rm = TRUE), min(Growth, na.rm = TRUE),
            min(CPI, na.rm = TRUE), min(Imports, na.rm = TRUE),
            min(Exports, na.rm = TRUE), min(Population, na.rm = TRUE)),
    Max = c(max(GDP, na.rm = TRUE), max(Growth, na.rm = TRUE),
            max(CPI, na.rm = TRUE), max(Imports, na.rm = TRUE),
            max(Exports, na.rm = TRUE), max(Population, na.rm = TRUE))
  ) %>%
  mutate(Variable = c("GDP", "Growth", "CPI", "Imports", "Exports", "Population")) %>%
  select(Variable, Mean, Median, SD, Min, Max)
```

## GDP, Exports, and Population Over Time

```{r plot-over-time, fig.width=8, fig.height=8}
p1 <- ggplot(df, aes(Year, GDP)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "GDP Over Time") +
  theme_minimal()

p2 <- ggplot(df, aes(Year, Population)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Population Over Time") +
  theme_minimal()

p3 <- ggplot(df, aes(Year, Exports)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Exports Over Time") +
  theme_minimal()

grid.arrange(p1, p2, p3)
```

# GDP Analysis

## Transformation and ACF/PACF

```{r gdp-transform}
gdp_transformed <- log(df$GDP)
```

```{r gdp-acf-pacf, fig.width=8, fig.height=6}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), cex = 1)
acf(df$GDP, lag.max = 20, main = "GDP ACF")
acf(gdp_transformed, lag.max = 20, main = "Log GDP ACF")
pacf(df$GDP, lag.max = 20, main = "GDP PACF")
pacf(gdp_transformed, lag.max = 20, main = "Log GDP PACF")
axis(1, at = 0:20, labels = 0:20)
```

## Model Fitting

```{r gdp-models}
m_ar1 <- Arima(gdp_transformed, order = c(1, 1, 0))
m_ma1 <- Arima(gdp_transformed, order = c(0, 1, 1))
m_arma <- Arima(gdp_transformed, order = c(1, 1, 1))

coeftest(m_ar1)
coeftest(m_ma1)
coeftest(m_arma)

summary(m_ar1)
summary(m_ma1)
summary(m_arma)
```

Fitted ARIMA models:

- AR(1): $\nabla y_t = 0.1642 \nabla y_{t-1} + \omega_t$
- MA(1): $\nabla y_t = \omega_t + 0.1267 \omega_{t-1}$
- ARMA(1,1): $\nabla y_t = 0.978521 \nabla y_{t-1} + \omega_t - 0.913256 \omega_{t-1}$

## Differenced Series Diagnostics

```{r gdp-diff-acf-pacf, fig.width=8, fig.height=5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), cex = 1)
acf(diff(gdp_transformed), lag.max = 20, main = "Differenced log GDP ACF")
pacf(diff(gdp_transformed), lag.max = 20, main = "Differenced log GDP PACF")
```

```{r gdp-sarima}
sarima(gdp_transformed, 1, 1, 1, no.constant = TRUE)
```

## Forecasting

```{r gdp-forecast}
gdp_ts <- ts(log(df$GDP), start = 1960, frequency = 1)
arima111 <- arima(gdp_ts, order = c(1, 1, 1))
f1 <- forecast(arima111)
f1

raw_gdp_ts <- ts(df$GDP, start = 1960, frequency = 1)
raw_arima111 <- arima(raw_gdp_ts, order = c(1, 1, 1))
raw_f1 <- forecast(raw_arima111)
raw_f1
```

```{r gdp-forecast-plot, fig.width=8, fig.height=6}
p1 <- autoplot(forecast(arima111)) +
  scale_x_continuous(breaks = pretty(time(gdp_ts), n = 10)) +
  labs(title = "ARIMA(1,1,1) - Transformed GDP")

p2 <- autoplot(forecast(raw_arima111)) +
  scale_x_continuous(breaks = pretty(time(raw_gdp_ts), n = 10)) +
  labs(title = "ARIMA(1,1,1) - Raw GDP")

grid.arrange(p1, p2)
```

# Exports Analysis

## Transformation and ACF/PACF

```{r exp-transform}
exp_transformed <- log(df$Exports)
```

```{r exp-acf-pacf, fig.width=8, fig.height=6}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), cex = 1)
acf(df$Exports, lag.max = 20, main = "Export ACF")
acf(exp_transformed, lag.max = 20, main = "Log Export ACF")
pacf(df$Exports, lag.max = 20, main = "Export PACF")
pacf(exp_transformed, lag.max = 20, main = "Log Export PACF")
axis(1, at = 0:20, labels = 0:20)
```

## Model Fitting

```{r exp-models}
e_ar1 <- Arima(exp_transformed, order = c(2, 1, 0))
e_ma1 <- Arima(exp_transformed, order = c(0, 1, 1))
e_arma <- Arima(exp_transformed, order = c(2, 1, 1))

coeftest(e_ar1)
coeftest(e_ma1)
coeftest(e_arma)

summary(e_ar1)
summary(e_ma1)
summary(e_arma)
```

Fitted ARIMA models:

- AR(2): $\nabla y_t = -0.2974 \nabla y_{t-1} + \omega_t$
- MA(1): $\nabla y_t = \omega_t - 0.3473 \omega_{t-1}$
- ARMA(2,1): $\nabla y_t = -0.0511 \nabla y_{t-1} + \omega_t - 0.3094 \omega_{t-1}$

## Differenced Series Diagnostics

```{r exp-diff-acf-pacf, fig.width=8, fig.height=5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), cex = 1)
acf(diff(exp_transformed), lag.max = 20, main = "Differenced log export ACF")
pacf(diff(exp_transformed), lag.max = 20, main = "Differenced log export PACF")
axis(1, at = 0:20)
```

```{r exp-sarima}
sarima(exp_transformed, 2, 1, 0, no.constant = TRUE)
```

## Forecasting

```{r exp-forecast}
exp_ts <- ts(log(df$Exports), start = 1960, frequency = 1)
e_arima210 <- arima(exp_ts, order = c(2, 1, 0))
f2 <- forecast(e_arima210)
f2

raw_exp_ts <- ts(df$Exports, start = 1960, frequency = 1)
raw_e_arima210 <- arima(raw_exp_ts, order = c(2, 1, 0))
raw_f2 <- forecast(raw_e_arima210)
raw_f2
```

```{r exp-forecast-plot, fig.width=8, fig.height=6}
p3 <- autoplot(forecast(e_arima210)) +
  scale_x_continuous(breaks = pretty(time(exp_ts), n = 10)) +
  labs(title = "ARIMA(2,1,0) - Transformed Exports")

p4 <- autoplot(forecast(raw_e_arima210)) +
  scale_x_continuous(breaks = pretty(time(raw_exp_ts), n = 10)) +
  labs(title = "ARIMA(2,1,0) - Raw Exports")

grid.arrange(p3, p4)
```

# Population Analysis

## Transformation and Initial Models

```{r pop-transform}
pop_transformed <- log(df$Population)
```

```{r pop-models}
p_ar1_pop <- Arima(pop_transformed, order = c(1, 1, 0))
p_ma1_pop <- Arima(pop_transformed, order = c(0, 1, 1))
diff_pop_transformed <- diff(pop_transformed)
p_arma_pop <- Arima(diff_pop_transformed, order = c(1, 0, 1))

coeftest(p_ar1_pop)
coeftest(p_ma1_pop)
coeftest(p_arma_pop)

summary(p_ar1_pop)
summary(p_ma1_pop)
summary(p_arma_pop)
```

Fitted ARIMA models (first-order differencing):

- AR(1): $\nabla y_t = 0.9954 \nabla y_{t-1} + \omega_t$
- MA(1): $\nabla y_t = \omega_t + 0.9999 \omega_{t-1}$
- ARMA(1,1): $\nabla y_t = 0.0181 + 0.9568 \nabla y_{t-1} + \omega_t + 0.9999 \omega_{t-1}$

## Differenced Series Diagnostics

```{r pop-diff-acf-pacf, fig.width=8, fig.height=5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), cex = 1)
acf(diff(pop_transformed), lag.max = 20, main = "Differenced log population ACF")
pacf(diff(pop_transformed), lag.max = 20, main = "Differenced log population PACF")
axis(1, at = 0:20)
```

```{r pop-sarima-1}
sarima(diff_pop_transformed, 1, 0, 1)
```

## Second-Order Differencing

```{r pop-models-2}
p_ar2_pop <- Arima(pop_transformed, order = c(1, 2, 0))
p_ma2_pop <- Arima(pop_transformed, order = c(0, 2, 1))
p_arma2_pop <- Arima(pop_transformed, order = c(1, 2, 1))

coeftest(p_ar2_pop)
coeftest(p_ma2_pop)
coeftest(p_arma2_pop)

summary(p_ar2_pop)
summary(p_ma2_pop)
summary(p_arma2_pop)
```

New fitted ARIMA models (second-order differencing):

- AR(1): $\nabla^2 y_t = 0.8932 \nabla^2 y_{t-1} + \omega_t$
- MA(1): $\nabla^2 y_t = \omega_t + 0.9994 \omega_{t-1}$
- ARMA(1,1): $\nabla^2 y_t = 0.8534 \nabla^2 y_{t-1} + \omega_t + 0.8062 \omega_{t-1}$

```{r pop-sarima-2}
sarima(pop_transformed, 1, 2, 1)
```

## Forecasting

```{r pop-forecast}
pop_ts <- ts(log(df$Population), start = 1960, frequency = 1)
arima121 <- arima(pop_ts, order = c(1, 2, 1))
f3 <- forecast(arima121)
f3

raw_pop_ts <- ts(df$Population, start = 1960, frequency = 1)
raw_arima121 <- arima(raw_pop_ts, order = c(1, 2, 1))
raw_f3 <- forecast(raw_arima121)
raw_f3
```

```{r pop-forecast-plot, fig.width=8, fig.height=6}
p5 <- autoplot(forecast(arima121)) +
  scale_x_continuous(breaks = pretty(time(pop_ts), n = 10)) +
  labs(title = "ARIMA(1,2,1) - Transformed Population")

p6 <- autoplot(forecast(raw_arima121)) +
  scale_x_continuous(breaks = pretty(time(raw_pop_ts), n = 10)) +
  labs(title = "ARIMA(1,2,1) - Raw Population")

grid.arrange(p5, p6)
```

# Cross-Correlation

```{r cross-correlation, fig.width=7, fig.height=5}
ccf(residuals(arima111), residuals(e_arima210), lag.max = 20,
    main = "Export(ARIMA(2,1,0)) and GDP(ARIMA(1,1,1)) Cross Correlation Function")
```
